---
title: "Surface Water"
output:
  html_document: default
  toc: true
  toc_depth: 3 
---

# The Prompt 

Background: The attached Excel file (Dataset_C) contains analytical chemistry data from a variety of surface water stations. 
 
Given the attached dataset:

* Develop a tool for plotting time series by station and analyte.

* Symbolize data by season.

* Add a horizontal line for the mean concentration to each plot.

* Which analytes have the most seasonal variability?  What tools/approach did you use?

* Produce a pdf of your plots.


# First Forays and Data Explorations

Before creating the final products, I've done some preparation and basic exploration, generally scoping out the data set, and preparing it for downstream analysis. 

```{r, setup, include=TRUE, warning=F, message=F}
rm(list=ls()) # Clear working environment

library("MapSuite") #Self-written library, has many common libs as dependencies
library("hexbin") # For the hex-bin plots

# Read in surface water observations as data.table sw
  sw<-fread("/Users/stubbsrw/Documents/git_code/stubbs_repo/fe_problems/data/Dataset_C.csv")

# Parse out date information from character date column
  sw[,index:=seq(1:nrow(sw))]
  sw[,Month:=as.numeric(strsplit(SampleDate,"/")[[1]][1]),by=index] 
  sw[,Day:=as.numeric(strsplit(SampleDate,"/")[[1]][2]),by=index]
  sw[,Year:=as.numeric(strsplit(SampleDate,"/")[[1]][3]),by=index]
  sw[,Date:=as.IDate(paste0((2000+Year),"-",Month,"-",Day))] # Create formal column of integer-date

# For each analyte, discover the minimum date, and generate a month index from that date:
  sw[,year_index:=Year-min(Year,na.rm=T),by=StandardAnalyte]
  sw[,year_min:=min(Year,na.rm=T),by=StandardAnalyte] # First sample year, calculated by each analyte
  sw[,month_index:=12*(Year-year_min) + Month] # Number of months from the start of the samples

# Add variables on mean concentration for each analyte by site and globally/for all samples
  sw[,mean_concentration_analyte_station:=mean(StandardResult,na.rm=T),by=list(StationName,StandardAnalyte)]
  sw[,mean_concentration_analyte:=mean(StandardResult,na.rm=T),by=StandardAnalyte]
```

## Assign seasons to different months

In order to symbolize by season, you first need to define what data points are in each season. For now, I'm 

```{r}
sw[Month %in% c(12,1,2), Season:='Winter']
sw[Month %in% c(3,4,5), Season:='Spring']
sw[Month %in% c(6,7,8), Season:='Summer']
sw[Month %in% c(9,10,11), Season:='Fall']
sw[,Month_String:=month.abb[Month]]
sw[,log_obs:=log(StandardResult)] # Just in case a log-transform would be more informative, although NaNs will exist where the observation is negative
```
Worth noting here that you could do this a fancier way if you wanted, based on cut points of the equinoxes; for now, we'll just go by month approximations. I'm also assuming that this is in the Northern hemisphere..

### Plotting Temperature over the full time series

```{r}
a<-"Temperature" #Picking this one since I have the best sense of what it "should" or "could" be per season
s<-"FE-4023" # arbitrary station to test things out on

# Make a subset data set with only the analyte you are interested in
analyte<-sw[StandardAnalyte==a]
mean_observation<-mean(analyte$StandardResult)

# Plotting using the HexBin Frequency graphics, where the number of stations with an observation in that category is essentially heat-mapped
ggplot(analyte, aes(x=Date, y=StandardResult)) + 
  geom_hex() + 
  xlab("Observations Over Time, All Stations") + 
  scale_x_date(labels = function(x) format(x, "%b-%y")) + 
  ylab(analyte$StandardUnit[1]) + 
  ggtitle(paste0(a)) + theme_bw() + 
  scale_fill_gradientn(colors=wpal("berries")) +
  guides(fill=guide_colourbar(title="N Stations", title.position="top", barheight=10, barwidth=1,
                              label=TRUE, ticks=FALSE, direction="vertical")) 
```

Let's try pooling observations across stations and also years, to get a rough sense of seasonality from a graphical persepective: 
```{r}
# Plotting by Month, for all years, all stations 
months_in_order<-c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")

# Non-Transformed
ggplot(analyte, aes(x= Month, y=log_obs)) + geom_hex() + 
  xlab("Observations in Each Month, All Years, All Stations") + 
  scale_x_continuous(limits=c(1,12),breaks=seq(1,12), labels=months_in_order) +
  ylab(analyte$StandardUnit[1]) + 
  ggtitle(paste0(a)) + theme_bw() + 
  scale_fill_gradientn(colors=wpal("berries")) +
  guides(fill=guide_colourbar(title="N Stations", title.position="top", barheight=10, barwidth=1,
                              label=TRUE, ticks=FALSE, direction="vertical")) 
```

## Plot by Station, Analyte. 

```{r}
for (s in unique(sw$StationName)){
for ( a in unique(sw$StandardAnalyte)){

}
}
```

## Fit GAMM with knots for each season and year

Another possible approach is to fit a general additive mixed model, where the long-term time trend and the short-term, periodic annual trends are conceptualized by cubic splines. 

```{r}
# Pulling from https://www.fromthebottomoftheheap.net/2014/05/09/modelling-seasonal-data-with-gam/
# 1 knot every 6 months, 1 knot for each year of data?
m0 <- mgcv::gamm(StandardResult ~ s(Month, bs = "cc", k = 6) + s(month_index, k = 3) + as.factor(StationName), data = analyte)

plot(m0$gam, scale = 0, main = "GAMM fit", xlab = "Month", ylab = paste0(a," Seasonal Effect"))

```

Let's try a linear model with a sin and cosine function

```{r}
mod <- lm(log_obs  ~ sin(2*pi*(month_index/12))+cos(2*pi*(month_index/12))+StationName,data=analyte)
summary(mod)

plot(log_obs~month_index,data=analyte)
plot(analyte$month_index,mod$fitted.values)
```
## Linear Model with 

```{r}
mod <- lm(StandardResult  ~ sin(2*pi*(month_index/12))+cos(2*pi*(month_index/12))+StationName+Year,data=analyte)
summary(mod)

plot(StandardResult~month_index,data=analyte)
lines(mod$fitted.values,analyte$month_index,col=2)

analyte[,fitted:=mod$fitted.values]
```

# Quantifying Seasonality

A few methods came to mind for this data set. 

For any given analyte, only a handful of observations were available across the time period for any given station-- which makes the variability in measurement due to station and due to other factors difficult to disentangle. Without knowing the spatial location of any of the stations, I assumed that the stations would be nearby one another, and subject to (at least roughly) the same weather and insolation patterns--- if these stations were far apart, it would be even more difficult to measure the "seasonal component" of each analyte, since the magnitude of seasonal changes could be subject to variables such as latitude and elevation.  

It seemed as if the ultimate goal of this question was to be able to the magnitude of seasonal variability across analytes, and 

Given these constraints, a few different strategies came to mind. For each of them, I tested the method on Temperature, since the right strategy would presumably show a seasonal effect for this analyte. 


```{r}
```